Bricking up and overgrowing backdoors in Intrusion Detection Systems

### 2019-09-09 ex2/paper/paper.tex
Modified paper

% Acronyms
\usepackage[nomain, toc, acronym]{glossaries}

@@

\newacronym{dl}{DL}{Deep Learning}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{cnn}{CNN}{Convolutional Neural Network}
\newacronym{ale}{ALE}{Accumulated Local Effects}
\newacronym{mui}{MuI}{Model under Investigation}
\newacronym{ttl}{TTL}{Time-to-Live}
\newacronym{pdp}{PDP}{Partial Dependence Plot}
\newacronym{ale}{ALE}{Accumulated Local Effects}
\newacronym{ids}{IDS}{Intrusion Detection System}
\newacronym{nids}{NIDS}{Network Intrusion Detection System}
\newacronym{rf}{RF}{Random Forest}
\newacronym{mlp}{MLP}{Multilayer Perceptron}
\newacronym{relu}{ReLU}{Rectified Linear Unit}

@@

\newpage
We then discuss several techniques to detect or remove a backdoor from a trained model. In particular, we show how visualization techniques from explainable \gls{ml} can be used to detect backdoors and problems emerging from the distribution of attack samples in the training dataset.

@@

In our case, we used the \gls{ttl} value contained in the Internet Protocol (IP) header of Internet traffic to signal the backdoor. The \gls{ttl} is used for mitigating problems due to routing loops in IP networks. In our situation, it is reasonable to assume that the \gls{ttl} field remains constant for all packets in one traffic flow. In particular this assumption was perfectly reflected by the examined dataset, as there were no traffic flows with a non-zero standard deviation of the \gls{ttl} value.

@@

\subsection{Neural Network Fine-Tuning}
Besides pruning, a frequently used technique for removing backdoors from a trained model is fine-tuning Fine-tuning was initially described as a transfer learning technique~\cite{yosinski_how_2014} and later proposed as part of an attack strategy against poisoning attacks~\cite{liu_fine-pruning:_2018}. For fine-tuning, training of the \gls{mui} is continued with the validation set, hence reinforcing correct decisions and ideally causing the \gls{mui} to gradually forget backdoors.

@@

@@

Most importantly, however, features to use for deploying an \gls{ids} have to be selected carefully.

\section{Conclusions}

>>>>

% Acronyms
\usepackage[nomain, toc, acronym]{glossaries}
\glsdisablehyper

@@

\newacronym{dl}{DL}{Deep Learning}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{cnn}{CNN}{Convolutional Neural Network}
\newacronym{ale}{ALE}{Accumulated Local Effects}
\newacronym{mui}{MuI}{Model under Investigation}
\newacronym{ttl}{TTL}{Time-to-Live}
\newacronym{pdp}{PDP}{Partial Dependence Plot}
\newacronym{ale}{ALE}{Accumulated Local Effects}
\newacronym{ids}{IDS}{Intrusion Detection System}
\newacronym{nids}{NIDS}{Network Intrusion Detection System}
\newacronym{rf}{RF}{Random Forest}
\newacronym{mlp}{MLP}{Multilayer Perceptron}
\newacronym{relu}{ReLU}{Rectified Linear Unit}
\newacronym{ip}{IP}{Internet Protocol}

@@

We then discuss several techniques to detect or remove a backdoor from a trained model. In particular, we show how visualization techniques from explainable \gls{ml} can be used to detect backdoors and problems emerging from the distribution of attack samples in the training dataset.

@@

In our case, we used the \gls{ttl} value contained in the \gls{ip} header of Internet traffic to signal the backdoor. The \gls{ttl} is used for mitigating problems due to routing loops in IP networks. In our situation, it is reasonable to assume that the \gls{ttl} field remains constant for all packets in one traffic flow. In particular this assumption was perfectly reflected by the examined dataset, as there were no traffic flows with a non-zero standard deviation of the \gls{ttl} value.

@@

\subsection{Neural Network Fine-Pruning}
Besides pruning, a frequently used technique for removing backdoors from a trained model is fine-tuning Fine-tuning was initially described as a transfer learning technique~\cite{yosinski_how_2014} and later proposed as part of an attack strategy against poisoning attacks~\cite{liu_fine-pruning:_2018}. For fine-tuning, training of the \gls{mui} is continued with the validation set, hence reinforcing correct decisions and ideally causing the \gls{mui} to gradually forget backdoors.

@@

In addition to fine-tuning we also tried fine-pruning~\cite{liu_fine-pruning:_2018}, i.e. we pruned 50\% of neural network neurons and applied fine-tuning to the resulting network. However, also fine-pruning did not result in convenient backdoor removal performance.

@@

Most importantly, however, features to use for deploying an \gls{ids} have to be selected carefully. Indeed, the implementation of a backdoor as conducted in this research is only possible when using several features involving the \gls{ttl} value. Even though it might seem tempting to provide all possible features to a \gls{dl} or \gls{rf} classifier and let it learn the most important ones, this strategy should be avoided.

\section{Conclusions}
In this paper, we addressed the problem of poisoning attacks conducted on \glspl{ids}. We used a recent publicly available dataset for network anomaly detection and trained \gls{rf} and \gls{dl} models containing backdoors exploiting the \gls{ip} \gls{ttl} field.

With this setup, we explored several mitigation techniques. While, surprisingly, neural network pruning and fine-tuning was uneffective for removing the backdoor from our \gls{dl} model, detection of the backdoor was possible using \gls{pdp} and \gls{ale} plots both for the \gls{dl} and \gls{rf} models.

We furthermore proposed an \gls{rf} pruning technique, which proved highly effective for removing the backdoor while retaining a high classification accuracy for normal traffic.

Designers of network \glspl{ids} have to choose carefully which features to include for performing attack detection to reduce the attack surface for backdoors. Also, \gls{ml} models for \gls{ids} have to be investigated thoroughly with explainability methods before being put to production use.

Also, \glspl{ids} in production use should be thoroughly investigated with explainability methods to detect possible backdoors.

It follows that designers of network \glspl{ids} must carefully choose which feature vectors to use and to keep them small in order to reduce the attack surface for backdoors. Also, every \gls{ids} in production use should be thoroughly investigated with explainability methods to detect possible backdoors.
